Reduced plasticity and variance in physiological rates of ectotherm populations under climate change

### 2024-03-13 ms.qmd
Update formatting for GCB

## **Materials and Methods**
### *Literature collection*
We compiled literature on ectothermic animals that measured physiological rates (e.g., metabolic rate) at two or more temperatures after having been acclimated (or acclimatized) at these temperatures. We used data from a previous meta-analysis [@Seebacher2015-wi] and updated Seebacher *et al.* @Seebacher2015-wi's data by extracting data from suitable studies from our own searches that followed the same search protocol. More specifically, we performed a literature search using the Web of Science database for articles or proceedings papers published in English from 2013 to 2017 (the date after Seebacher *et al.* [@Seebacher2015-wi] searches were conducted) using the following topic search string: *"(acclimat* AND (therm* OR temp*) NOT (plant* OR tree* OR forest* OR fung* OR mammal* OR marsup* OR bird* OR human OR exercis* OR train* OR hypoxi*))"*. We further limited to the following research areas: Anatomy Morphology; Biodiversity Conservation; Biology; Ecology; Endocrinology Metabolism; Entomology; Evolutionary Biology; Marine Freshwater Biology; Physiology; Respiratory System, Reproductive Biology, Zoology.

Our search resulted in 1,321 papers for screening in Rayyan [@Ouzzani2016-ge]. We also cross-checked papers we found in our searches with a recent paper by Havird *et al.* @Havird2020-mj, which also updates the dataset of Seebacher *et al.* @Seebacher2015-wi's. We included any papers that were missed between our searches and those of Havird *et al.* @Havird2020-mj. Havird *et al.* @Havird2020-mj added 7 new studies (mainly because they were focused on metabolic rates), and our searches differed from theirs by only a single paper [i.e., @Bulgarella2015-yl]. Given the physiological traits we included were broader, we had a substantial increase in additional papers that we added to Seebacher *et al.* @Seebacher2015-wi's dataset. More specifically, in addition to the `r length(unique(old_papers$record_num))` papers we included from the Seebacher *et al.* @Seebacher2015-wi dataset, we extracted data from an extra `r length(unique(new_papers$record_num))` papers (with a total of `r dim(new_papers)[1]` effects; a `r  (length(unique(new_papers$record_num)) / length(unique(old_papers$record_num)))*100`% increase in the number of published articles). Note that Seebacher *et al.* @Seebacher2015-wi included a total of 205 publications, however, not all these contained the necessary statistics we needed to derive effect sizes and associated sampling variances (see below). While we may have missed papers, our goal was to obtain a large representative (and unbiased) sample of acclimation research rather than a comprehensive dataset. As such, our database represents the most up-to-date dataset used by Seebacher *et al.* @Seebacher2015-wi to answer questions on physiological rates across ectotherms.

@@

Following Noble *et al.* @Noble2022-ty we calculated a series of temperature-corrected effect sizes that compared mean physiological rates ($lnRR_{Q_{10}}$) as well as the variability in physiological rates ($lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$). These effect sizes are similar to the traditional temperature coefficient ($Q_{10}$), but with formal analytical approximations of their sampling variances. Sampling variances for effect sizes allowed us to make use of traditional meta-analytic modelling approaches.

@@

#### *Comparing variance in physiological rates*
Nakagawa *et al.* @Nakagawa2015-fo proposed analogous effect size estimates to *lnRR* that allow for comparisons of changes in variance between two groups, the log variance ratio (*lnVR*) and the log coefficient of variation (*lnCVR*). *lnVR* and *lnCVR* are ratios that describe the relative difference in trait variability between two groups. We refer readers to Nakagawa *et al.* @Nakagawa2015-fo for the equations describing *lnVR* and *lnCVR*, but these can easily be extended to their $Q_{10}$ analogues (and associated sampling variance) as follows:

@@

 @eq-lnq10VR and @eq-slnq10VR describe the change in physiological rate variance (@eq-lnq10VR) across a 10&deg;C temperature change along with its sampling variance (@eq-slnq10VR). While this is a useful metric, as discussed by Nakagawa *et al.* @Nakagawa2015-fo there is often a strong mean-variance relationship that needs to be accounted for in analysing changes in variance. As such, we calculated the coefficient of variation, which standardizes changes in variance for changes in means as follows:

@@

### ***Plasticity and variance in physiological rates do not differ between life-stages***
Life-history stages often occupy different ecological niches and exhibit different physiological responses, levels of plasticity, and patterns of mortality in response to temperature. As such, it is becoming increasingly important to understand how climate change will impact different life-history stages [@levy2015resolving; @petitgas2013impacts]. Greater plasticity in early life-stages of development is expected to be important to increase resilience to the effects of climate change because early life-stages are often particularly vulnerable periods in development [@martin2015age; @stearns1976life]. We show that there were generally similar patterns between early and late life stages across a diversity of taxa, both in terms of variance changes and the capacity for plasticity. On average, our results show that early life stages are not more vulnerable to the impacts of climate change. However, our general meta-analytic findings do not imply that early life stages are not always less plastic, but that such responses are likely context or trait depedent [@carter2020life; @moghadam2019heat]. For example, Moghadam *et al.* @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, Carter *et al.* @carter2020life showed greater thermal plasticity in metabolism for adults but little in pupae of *Onthophagus taurus*.  Despite variation in the literature, very few studies measure multiple life-history stages within the same population. In future research, there needs to be more studies that explicitly compare differences in plasticity among life stages across multiple traits within the same population.

### ***Ecological consequences of reduced variation in physiological rates***
Variance reductions in metabolism and performance (e.g., sprint speed) were most strongly impacted by increased temperature. Changes in variability of traits governing energy demand could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key; @seebacher2023can; @barneche2021warming]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, Kooijman *et al.* @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. Our results suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity in terrestrial ecosystems. However, data testing the impacts of changes in variance on populations are limited to very few taxa. Future experimentation across a broader range of taxa should manipulate levels of variance to understand its ecological outcomes.

@@

Theoretical evolutionary models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly when oscillations are predictable over time to make environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely to evolve in terrestrial environments. However, we did not find evidence to this effect. Our finding is consistent with other meta-analyses [@barley2021limited; @Gunderson2015-hz] with thermal tolerance (which we did not consider here). Modelling correct microenvironments for over 200 species across such diverse habitats is challenging. The lack of a relationship between physiological plasticity and environmental variability and predictability could be due the fact that ectotherms simply seek out microhabitats that make their environment quite stable [@huey2012predicting].  Therefore, it is possible that more fine-scale resolution of the habitats that organisms occupy will elucidate expected patterns. Randomly fluctuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), Hallsson *et al.* @hallsson2012selection showed that experimentally evolving populations under randomly fluctuating thermal conditions did not show any increase in plasticity, but had reduced plasticity. Leung *et al.* @Leung2020 also showed reduced plasticity in morphological traits of experimentally evolving algae populations (*Dunaliella salina*) when environments were less predictable [see also @leung2023phenotypic]. While it is unclear how these findings relate directly to acclimation because they integrate multiple forms of plasticity, they do suggest that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticity when environments are not predictable [@Leung2020; @hendry2016key]. The fact that aquatic ectotherms did show a capacity to acclimate suggests that environmental predictability, which is expected to be higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity.

>>>>

## **Materials and Methods**
### *Literature collection*
We compiled literature on ectothermic animals that measured physiological rates (e.g., metabolic rate) at two or more temperatures after having been acclimated (or acclimatized) at these temperatures. We used data from a previous meta-analysis [@Seebacher2015-wi] and updated @Seebacher2015-wi's data by extracting data from suitable studies from our own searches that followed the same search protocol. More specifically, we performed a literature search using the Web of Science database for articles or proceedings papers published in English from 2013 to 2017 (the date after [@Seebacher2015-wi] searches were conducted) using the following topic search string: *"(acclimat* AND (therm* OR temp*) NOT (plant* OR tree* OR forest* OR fung* OR mammal* OR marsup* OR bird* OR human OR exercis* OR train* OR hypoxi*))"*. We further limited to the following research areas: Anatomy Morphology; Biodiversity Conservation; Biology; Ecology; Endocrinology Metabolism; Entomology; Evolutionary Biology; Marine Freshwater Biology; Physiology; Respiratory System, Reproductive Biology, Zoology.

Our search resulted in 1,321 papers for screening in Rayyan [@Ouzzani2016-ge]. We also cross-checked papers we found in our searches with a recent paper by @Havird2020-mj, which also updates the dataset of @Seebacher2015-wi's. We included any papers that were missed between our searches and those of @Havird2020-mj. @Havird2020-mj added 7 new studies (mainly because they were focused on metabolic rates), and our searches differed from theirs by only a single paper [i.e., @Bulgarella2015-yl]. Given the physiological traits we included were broader, we had a substantial increase in additional papers that we added to @Seebacher2015-wi's dataset. More specifically, in addition to the `r length(unique(old_papers$record_num))` papers we included from the @Seebacher2015-wi dataset, we extracted data from an extra `r length(unique(new_papers$record_num))` papers (with a total of `r dim(new_papers)[1]` effects; a `r  (length(unique(new_papers$record_num)) / length(unique(old_papers$record_num)))*100`% increase in the number of published articles). Note that @Seebacher2015-wi included a total of 205 publications, however, not all these contained the necessary statistics we needed to derive effect sizes and associated sampling variances (see below). While we may have missed papers, our goal was to obtain a large representative (and unbiased) sample of acclimation research rather than a comprehensive dataset. As such, our database represents the most up-to-date dataset used by @Seebacher2015-wi to answer questions on physiological rates across ectotherms.

@@

Following @Noble2022-ty we calculated a series of temperature-corrected effect sizes that compared mean physiological rates ($lnRR_{Q_{10}}$) as well as the variability in physiological rates ($lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$). These effect sizes are similar to the traditional temperature coefficient ($Q_{10}$), but with formal analytical approximations of their sampling variances. Sampling variances for effect sizes allowed us to make use of traditional meta-analytic modelling approaches.

@@

#### *Comparing variance in physiological rates*
@Nakagawa2015-fo proposed analogous effect size estimates to *lnRR* that allow for comparisons of changes in variance between two groups, the log variance ratio (*lnVR*) and the log coefficient of variation (*lnCVR*). *lnVR* and *lnCVR* are ratios that describe the relative difference in trait variability between two groups. We refer readers to @Nakagawa2015-fo for the equations describing *lnVR* and *lnCVR*, but these can easily be extended to their $Q_{10}$ analogues (and associated sampling variance) as follows:

@@

 @eq-lnq10VR and @eq-slnq10VR describe the change in physiological rate variance (@eq-lnq10VR) across a 10&deg;C temperature change along with its sampling variance (@eq-slnq10VR). While this is a useful metric, as discussed by @Nakagawa2015-fo there is often a strong mean-variance relationship that needs to be accounted for in analysing changes in variance. As such, we calculated the coefficient of variation, which standardizes changes in variance for changes in means as follows:

@@

### ***Plasticity and variance in physiological rates do not differ between life-stages***
Life-history stages often occupy different ecological niches and exhibit different physiological responses, levels of plasticity, and patterns of mortality in response to temperature. As such, it is becoming increasingly important to understand how climate change will impact different life-history stages [@levy2015resolving; @petitgas2013impacts]. Greater plasticity in early life-stages of development is expected to be important to increase resilience to the effects of climate change because early life-stages are often particularly vulnerable periods in development [@martin2015age; @stearns1976life]. We show that there were generally similar patterns between early and late life stages across a diversity of taxa, both in terms of variance changes and the capacity for plasticity. On average, our results show that early life stages are not more vulnerable to the impacts of climate change. However, our general meta-analytic findings do not imply that early life stages are not always less plastic, but that such responses are likely context or trait depedent [@carter2020life; @moghadam2019heat]. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, @carter2020life showed greater thermal plasticity in metabolism for adults but little in pupae of *Onthophagus taurus*.  Despite variation in the literature, very few studies measure multiple life-history stages within the same population. In future research, there needs to be more studies that explicitly compare differences in plasticity among life stages across multiple traits within the same population.

### ***Ecological consequences of reduced variation in physiological rates***
Variance reductions in metabolism and performance (e.g., sprint speed) were most strongly impacted by increased temperature. Changes in variability of traits governing energy demand could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key; @seebacher2023can; @barneche2021warming]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. Our results suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity in terrestrial ecosystems. However, data testing the impacts of changes in variance on populations are limited to very few taxa. Future experimentation across a broader range of taxa should manipulate levels of variance to understand its ecological outcomes.

@@

Theoretical evolutionary models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly when oscillations are predictable over time to make environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely to evolve in terrestrial environments. However, we did not find evidence to this effect. Our finding is consistent with other meta-analyses [@barley2021limited; @Gunderson2015-hz] with thermal tolerance (which we did not consider here). Modelling correct microenvironments for over 200 species across such diverse habitats is challenging. The lack of a relationship between physiological plasticity and environmental variability and predictability could be due the fact that ectotherms simply seek out microhabitats that make their environment quite stable [@huey2012predicting].  Therefore, it is possible that more fine-scale resolution of the habitats that organisms occupy will elucidate expected patterns. Randomly fluctuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), @hallsson2012selection showed that experimentally evolving populations under randomly fluctuating thermal conditions did not show any increase in plasticity, but had reduced plasticity. @Leung2020 also showed reduced plasticity in morphological traits of experimentally evolving algae populations (*Dunaliella salina*) when environments were less predictable [see also @leung2023phenotypic]. While it is unclear how these findings relate directly to acclimation because they integrate multiple forms of plasticity, they do suggest that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticity when environments are not predictable [@Leung2020; @hendry2016key]. The fact that aquatic ectotherms did show a capacity to acclimate suggests that environmental predictability, which is expected to be higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity.
